Lawvere-Tierney Sheafification\\ in Homotopy Type Theory

### 2015-01-13 paper/sheaf_lics.tex
add .bib
beginning of the introduction

  Sheafification is a popular tool in (higher) topos theory which
  allows to extend the internal logic of a topos with new
  principles. One of its most famous applications is the possibility
  to transform a topos into a boolean (classical) topos using the
  dense topology, which corresponds in essence to Gödel's double
  negation translation.
  % 
  The same construction has not been considered in Martin-Löf type
  theory because of a mismatch between topos theory and type theory. This
  mismatched has been fixed recently by considering homotopy type
  theory, an extension of Martin-Löf type theory with new
  principles inspired by category theory and homotopy theory, and
  which corresponds closely to higher toposes.
  % 
  In this paper, we give a computer-checked construction of
  Lawvere-Tierney sheafification in homotopy type theory, which allows
  in particular to give a meaning to the (propositional) law of
  excluded middle inside homotopy type theory, without axiom, being
  compatible with the full type-theoretic axiom of choice.

@@

\begin{itemize}
\item 
  links between HoTT and higher topos theory 
\item 
  modalities in HoTT
\item 
  Giraud's axiom
\item
  previous attempts to extend Gödel's double negation to 
  Martin-Löf type theory
\item 
  compare to the work of Parigot on $\lambda\mu$-calculus (Classical proofs as programs), and Coquand 
  (A realization of the negative interpretation of the Axiom of
  Choice)
\item
  work of Hugo Herbelin, A Constructive Proof of Dependent Choice,
  Compatible with Classical Logic
\item
  Barthe (CPS Translating Inductive and Coinductive Types)
  ``No translation is possible along the same lines for small $\Sigma$-types
  and sum types with dependent case.''
\item

\end{itemize}

@@

  $$\xymatrix{
    \sum_{e:E} (\chi~ e) \ar[r]^f \ar[d]_{\pi_1} & F \\
    E \ar@{-->}[ru]&
  }$$
\end{defi}

@@

  $$\xymatrix{
    \sum_{e:E} (\chi~ e) \ar[r]^f \ar[d]_{\pi_1} & F \\
    E \ar@{-->}[ru]&
  }$$
\end{defi}

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\begin{thebibliography}{4}

\bibitem[RS13]{sets_in_hott}
Egbert Rijke and Bas Spitters.
\newblock Sets in homotopy type theory.
\newblock 2013.

\bibitem[{Uni}13]{hottbook}
{Univalent Foundations Project}.
\newblock {\em Homotopy Type Theory: Univalent Foundations for Mathematics}.
\newblock 2013.

\bibitem[MM92]{maclanemoerdijk}
Saunders MacLane and Ieke Moerdijk.
\newblock {\em Sheaves in Geometry and Logic}.
\newblock Springer-Verlag, 1992.

\bibitem[HTT]{lurie}
Jacob Lurie.
\newblock{\em Higher Topos Theory}.

\bibitem[AKL]{lumsdaine}
Peter LeFanu Lumsdaine, Jeremy Avigad and Chris Kapulkin.
\newblock{\em Homotopy limits in type theory}.

\end{thebibliography}

>>>>

  Sheafification is a popular tool in (higher) topos theory which
  allows to extend the internal logic of a topos with new
  principles. One of its most famous applications is the possibility
  to transform a topos into a boolean (classical) topos using the
  dense topology, which corresponds in essence to Gödel's double
  negation translation.
  % 
  The same construction has not been considered in Martin-Löf type
  theory because of a mismatch between topos theory and type theory. This
  mismatched has been fixed recently by considering homotopy type
  theory, an extension of Martin-Löf type theory with new
  principles inspired by category theory and homotopy theory, and
  which corresponds closely to higher topoi.
  % 
  In this paper, we give a computer-checked construction of
  Lawvere-Tierney sheafification in homotopy type theory, which allows
  in particular to give a meaning to the (propositional) law of
  excluded middle inside homotopy type theory, without axiom, being
  compatible with the full type-theoretic axiom of choice.

@@

The sheaf construction, or sheafification~\cite{maclanemoerdijk}, is a
very powerful tool to construct new topoi from already existing ones,
allowing logical principles---that can not be proven to be true or false
in the old topos---to be valided (or invalided) in the new topos.
%
The most common use of sheafification is to construct a boolean
topos, thus validating the classical principle of the law of excluded
middle (EM), from an intuitionistic topos. 
%
An other famous (revisited) application has been developed by Paul
Cohen~\cite{cohen1966} to prove that the continuum hypothesis is
independent of the usual axioms of Zermelo-Fr\ae nkel set theory, even
in presence of the axiom of choice (AC).

Even if similar questions have been considered around the Curry-Howard
isomorphism, there is no canonical way to extend a programming language
such as type theory while keeping consistency automatically. 
%
For instance, much effort have been done to provide a computational
content to the law of excluded middle in order to define a constructive
version of classical logic. This has lead to various calculi, with
most notably the $\lambda \mu$-calculus of
Parigot~\cite{parigot1993classical}, but this has not allow to define
a new version of type theory with classical principles.
%
Other works have tried to extend continuation-passing-style (CPS)
transformation to type theory, but they have been faced with the
difficulty that the CPS transformation is incompatible with (full) dependent
sums~\cite{barthe2002cps}, which puts emphasis on the tedious link
between the axiom of choice and the law of excluded middle in type theory.

Nevertheless the axiom of choice Coquand, Herbelin, Krivine.
%
% Some works have tried to provide a realization of the axiom of choice 
% compatible with classical logic

\begin{itemize}
\item 
  links between HoTT and higher topos theory 
\item 
  modalities in HoTT
\item 
  Giraud's axiom
\end{itemize}

@@

  $$\xymatrix{
    \sum_{e:E} (\chi~ e) \ar[r]^f \ar[d]_{\pi_1} & F \\
    E \ar@{-->}[ru]_{!}&
  }$$
\end{defi}

@@

  $$\xymatrix{
    \sum_{e:E} (\chi~ e) \ar[r]^f \ar[d]_{\pi_1} & F \\
    E \ar@{-->}[ru]_{\exists !}&
  }$$
\end{defi}

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliographystyle{IEEEtran}
\bibliography{sheaf_lics}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
% \begin{thebibliography}{5}

% \bibitem[RS13]{sets_in_hott}
% Egbert Rijke and Bas Spitters.
% \newblock Sets in homotopy type theory.
% \newblock 2013.

% \bibitem[{Uni}13]{hottbook}
% {Univalent Foundations Project}.
% \newblock {\em Homotopy Type Theory: Univalent Foundations for Mathematics}.
% \newblock 2013.

% \bibitem[MM92]{maclanemoerdijk}
% Saunders MacLane and Ieke Moerdijk.
% \newblock {\em Sheaves in Geometry and Logic}.
% \newblock Springer-Verlag, 1992.

% \bibitem[HTT]{lurie}
% Jacob Lurie.
% \newblock{\em Higher Topos Theory}.

% \bibitem[AKL]{lumsdaine}
% Peter LeFanu Lumsdaine, Jeremy Avigad and Chris Kapulkin.
% \newblock{\em Homotopy limits in type theory}.

% \end{thebibliography}
